Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-04-05 pkg/disttree/inst/draft/draft.Rnw
small corrections and additional reference for Generalized Random Forests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2126 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 14070a331da16bc471dee29f4ce6ae11c3a72ec3

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero (i.e., dry days without
rain), zero-censored Gaussian distribution with mean/location parameter
$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
$$
\ell(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
\end{cases}
$$
where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
Other distributions $\mathcal{D}$ and corresponding log-likelihoods
$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.

With the specification of the distribution family and its log-likelihood 
function the task of fitting a distributional model turns into the task 
of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
maximum likelihood (ML) based on the learning sample
$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
$$ \label{eq:mle}
\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
$$
Equivalently, this can be defined based on the corresponding first-order
conditions
$$ \label{eq:foc}
\sum_{i = 1}^n s(\bm{\hat \theta}, y_i) = 0,
$$
where $s(\bm{\theta}, y_i)$ is the associated score function
$$ \label{eq:score}
s(\bm{\theta}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
$$ 
The latter is subsequently employed as a general goodness-of-fit measure
to assess how well the distribution with parameters $\bm{\theta}$ fits
one individual observation $y_i$.

@@

@@

To remove large parts of the skewness of precipitation data, 
a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
However, the power parameter may vary for different climatic zones or temporal 
aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
in their choice of $\frac{1}{1.6}$ as a suitable power parameter for the
region of Tyrol. The same power transformation is applied to both the
observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
``convective available potential energy'' (\emph{cape}).

@@

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
  the most relevant predictors based on meteorological expert knowledge
  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
  on the 80~available variables, 8 terms are included in the location
  submodel and 6 in the scale submodel. Both involve an interaction of
  \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
  to capture the potential for thunderstorms that frequently occur in
  summer afternoons in the Alps. The model is estimated by maximum
  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

@@

To assess the predictive performance, a full cross-validation is carried out
rather than relying on just the one fixed test set for the years 2009--2012.
To do so, a 10 times 7-fold cross-validation is carried out where each
splits the available 28~years into 7~subsets of 4~randomly selected years.
The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
(= 4~years) using the average CRPS across all observations.
The resulting 10~CRPS skill scores are displayed by boxplots in
Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4~percent, the distributional forest has a slightly higher improvement
of around 5.5~percent in median.

@@

The fitted distributional forest for July 24 and
observation station Axams illustrated in Figure~\ref{fig:axams-dist} 
is reproducible using the \textsf{R} code provided as demo in the package
\textbf{disttree}. This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
PIT histograms (Figure~\ref{fig:axams-pit})
and QQR plots (Figure~\ref{fig:axams-qq}).
Full replication of all results can be obtained using the companion 
\textsf{R} package \textbf{RainTyrol} and the \textsf{R} code of
\code{replication_code.R} provided in the package \textbf{disttree}.

@@

After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
the set of observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
Section~\ref{sec:distfit} the resulting estimated parameter 
$\bm{\hat{\theta}} = 
(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
can be plugged in the score function $s(\bm{\theta}, Y)$.
In that way a measurement of goodness-of-fit is obtained for each
parameter $\theta_j$ and each observation $y_i$.\\
To use this information, statistical tests are employed to detect
dependencies between the score values
$$
s(\bm{\hat{\theta}}, y) = 
%\begin{pmatrix} 
%s(\bm{\hat{\theta}}, y_1) \\
%s(\bm{\hat{\theta}}, y_2) \\
%\vdots \\
%s(\bm{\hat{\theta}}, y_n)
%\end{pmatrix} =
\begin{pmatrix} 
s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
s(\bm{\hat{\theta}}, y_2)_1 & s(\bm{\hat{\theta}}, y_2)_2 & \ldots & s(\bm{\hat{\theta}}, y_2)_k\\
\vdots & \vdots & \ddots & \vdots \\
s(\bm{\hat{\theta}}, y_n)_1 & s(\bm{\hat{\theta}}, y_n)_2 & \ldots & s(\bm{\hat{\theta}}, y_n)_k
\end{pmatrix}
$$
and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
\begin{align*}
H_0^l:  s(\bm{\hat{\theta}}, y) \qquad \bot \qquad Z_l 
\end{align*}

@@

With the conditional expectation $\mu_{l}$ and the covariance 
$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
statistic can be standardized. The observed multivariate linear 
statistic $t$ is mapped into the real line by a  univariate test 
statistic $c$. In the application of this paper a quadratic form 
is chosen, such that
$$
c_{quad}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
$$
where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
the maximum of the absolute values of the standardized linear statistic
can be considered.

The smaller the p-value corresponding to the standardized test 
statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_l$.

After Bonferroni-adjusting the p-values it has to be assessed whether
any of the resulting p-values is beneath the selected significance level. 
If so, the partitioning variable $Z_{l^\ast}$ with the lowest p-value 
is chosen as splitting variable.

As split point the breakpoint that leads to the highest discrepancy 
between score functions in the two resulting subgroups is selected. 
This difference can be measured by the linear statistic
$$
T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
$$
where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
The split point is then chosen as follows:
$$
r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
$$

Repeat the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal p-value for the statistical tests. In that way pre-pruning
is applied in order to find right-sized trees and hence 
avoid overfitting.

@@

As an alternative to QQ~plots based on randomized quantile residuals,
probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
histograms are a commonly used tool to assess how well-calibrated distributional 
fits are. Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
some deviations the overall calibration of all models is acceptable.

@@

@
%\end{subfigure}
\caption{\label{fig:pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
\end{figure}

>>>>

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero (i.e., dry days without
rain), zero-censored Gaussian distribution with mean/location parameter
$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
$$
\ell(\mu, \sigma; Y) = 
\begin{cases}
    %\log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    %\log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
    \log\left\{\sigma^{-1} \cdot \phi\left((Y - \mu) \cdot \sigma^{-1}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left((-\mu) \cdto \sigma^{-1}\right)\right\}, & \text{if } Y = 0
\end{cases}
$$

where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
Other distributions $\mathcal{D}$ and corresponding log-likelihoods
$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.

With the specification of the distribution family and its log-likelihood 
function the task of fitting a distributional model turns into the task 
of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
maximum likelihood (ML) based on the learning sample
$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
$$ \label{eq:mle}
\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
$$
Equivalently, this can be defined based on the corresponding first-order
conditions
$$ \label{eq:foc}
\sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
$$
where $s(\bm{\hat \theta}, y_i)$ is the associated score function
$$ \label{eq:score}
s(\bm{\hat{\theta}}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i) \mid_{\bm{\theta} = \bm{\hat{\theta}}}.
$$ 
The latter is subsequently employed as a general goodness-of-fit measure
to assess how well the distribution with parameters $\bm{\theta}$ fits
one individual observation $y_i$.

@@

Distributional random forests share some concepts and algorithmic aspects 
with other generalizations of Breiman and Cutler's random forests. Nearest 
neighbor weights are employed for aggregation in survival forests 
\citep{Hothorn+Lausen+Benner:2004},
quantile regression forests \citep{Meinshausen:2006},
transformation forests \citep{Hothorn+Zeileis:2017},
and generalized random forests for causal inferences 
\citep{Athey+Tibshirani+Wager:2017}.
These procedures aggregate over trees fitted to specific score functions 
(e.g., log rank scores in survival trees, model residuals in transformation 
or generalized forests). Distributional random forests, in contrast to these 
nonparametric approaches, provide a compromise between model flexibility 
and interpretability: The parameters of a problem-specific distribution 
(zero left-censored normal for precipitation) have a clear meaning but 
may depend on external variables in a quite general way.

@@

To remove large parts of the skewness of precipitation data, 
a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
However, the power parameter may vary for different climatic zones or temporal 
aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
in their choice of $1.6^{-1}}$ as a suitable power parameter for the
region of Tyrol. The same power transformation is applied to both the
observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
``convective available potential energy'' (\emph{cape}).

@@

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
  the most relevant predictors based on meteorological expert knowledge
  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
  on the 80~available variables, eight terms are included in the location
  submodel and six in the scale submodel. Both involve an interaction of
  \emph{tp} and \emph{cape} in the afternoon (between 1200\,UTC and 1800\,UTC)
  to capture the potential for thunderstorms that frequently occur in
  summer afternoons in the Alps. The model is estimated by maximum
  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

@@

To assess the predictive performance, a full cross-validation is carried out
rather than relying on just the one fixed test set for the years 2009--2012.
To do so, a 10 times 7-fold cross-validation is carried out where each
splits the available 28~years into 7~subsets of 4~randomly selected years.
The models are learned on six~folds (= 24~years) and evaluated on the 7-th fold
(= 4~years) using the average CRPS across all observations.
The resulting ten~CRPS skill scores are displayed by boxplots in
Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
line at a CRPSS of 0). Both GAMLSS models and the distributional forest 
perform distinctly better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4~percent, the distributional forest has a slightly higher improvement
of around 5.5~percent in median.

@@

The fitted distributional forest for July 24 and
observation station Axams illustrated in Figure~\ref{fig:axams-dist} 
is reproducible using the \textsf{R} code provided as demo in the package
\textbf{disttree}. This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
PIT histograms (Figure~\ref{fig:axams-pit})
and QQR plots (Figure~\ref{fig:axams-qq}).
Full replication of all results can be obtained using the companion 
\textsf{R} package \textbf{RainTyrol} and the \textsf{R} code of
\code{replication.R} provided in the package \textbf{disttree}.

@@

After fitting a distributional model $\mathcal{D}(Y, \bm{\theta})$ to 
the set of observations $\{y_i\}_{i = 1,\ldots,n}$ as explained in 
Section~\ref{sec:distfit} the resulting estimated parameter 
$\bm{\hat{\theta}} = 
(\hat{\theta}_1, \ldots, \hat{\theta}_k)$, $k \in \mathbb{N}$ 
can be plugged in the score function $s(\bm{\hat{\theta}}, Y)$.
In that way a measurement of goodness-of-fit is obtained for each
parameter $\theta_j$ and each observation $y_i$.\\
To use this information, statistical tests are employed to detect
dependencies between the score values
$$
s(\bm{\hat{\theta}}, y) = 
%\begin{pmatrix} 
%s(\bm{\hat{\theta}}, y_1) \\
%s(\bm{\hat{\theta}}, y_2) \\
%\vdots \\
%s(\bm{\hat{\theta}}, y_n)
%\end{pmatrix} =
\begin{pmatrix} 
s(\bm{\hat{\theta}}, y_1)_1 & s(\bm{\hat{\theta}}, y_1)_2 & \ldots & s(\bm{\hat{\theta}}, y_1)_k\\
s(\bm{\hat{\theta}}, y_2)_1 & s(\bm{\hat{\theta}}, y_2)_2 & \ldots & s(\bm{\hat{\theta}}, y_2)_k\\
\vdots & \vdots & \ddots & \vdots \\
s(\bm{\hat{\theta}}, y_n)_1 & s(\bm{\hat{\theta}}, y_n)_2 & \ldots & s(\bm{\hat{\theta}}, y_n)_k
\end{pmatrix}
$$
and each variable $Z_l \in \{Z_1, \ldots, Z_m\}$.  
\begin{align*}
H_0^l:  s(\bm{\hat{\theta}}, Y) \qquad \bot \qquad Z_l 
\end{align*}

@@

With the conditional expectation $\mu_{l}$ and the covariance 
$\Sigma_{l}$ as derived by \cite{Strasser+Weber:1999} the test 
statistic can be standardized. The observed multivariate linear 
statistic $t$ is mapped onto the real line by a  univariate test 
statistic $c$. In the application of this paper a quadratic form 
is chosen, such that
$$
c_{\text{quad}}(t,\mu,\Sigma) = (t-\mu)\Sigma^+(t-\mu)^{\top}
$$
where $\Sigma^+$ is the Moore-Penrose inverse of $\Sigma$. Alternatively,
the maximum of the absolute values of the standardized linear statistic
can be considered.

The smaller the $p$-value corresponding to the standardized test 
statistic $c(t_{l},\mu_{l},\Sigma_{l})$ is the stronger the discrepancy 
from the assumption of independence between the scores and the split 
variable $Z_l$.

After Bonferroni-adjusting the $p$-values it has to be assessed whether
any of the resulting $p$-values is beneath the selected significance level. 
If so, the partitioning variable $Z_{l^\ast}$ with the lowest $p$-value 
is chosen as splitting variable.

The breakpoint that leads to the highest discrepancy between score functions 
in the two resulting subgroups is selected as split point. 
This difference can be measured by the linear statistic
$$
T_{l^{\ast}}^r = \sum_{i \in \mathcal{B}_{1r}} s(\bm{\hat{\theta}}, y_i)
$$
where $\mathcal{B}_{1r}$ is the first of the two new subgroups that 
are defined by splitting in split point $r$ of variable $Z_{l^{\ast}}$. 
The split point is then chosen as follows:
$$
r^{\ast} = \argmin_{r} c(t_{l^{\ast}}^r,\mu_{j^{\ast}}^r,\Sigma_{l^{\ast}}^r).
$$

One repeats the testing and splitting procedure in each of the resulting 
subgroups until some stopping criterion is reached. This criterion 
can for example be a minimal number of observations in a node or a 
minimal $p$-value for the statistical tests. In that way pre-pruning
is applied in order to find right-sized trees and hence 
avoid overfitting.

@@

As an alternative to QQ~plots based on randomized quantile residuals,
probability integral transform (PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) 
histograms are a commonly used tool to assess how well-calibrated distributional 
fits are. Figure~\ref{fig:axams-pit} shows out-of-sample PIT histograms for the same
probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that despite
some deviations the overall calibration of all models is acceptable.

@@

@
%\end{subfigure}
\caption{\label{fig:axams-pit}Out-of-sample PIT histograms (2009--2012) for station Axams and models learned on data from 1985--2008.}
\end{figure}
